---
title: "Southern Ocean Krill Biodiversity Analysis"
author: "Pablo Deschepper"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    browser: true
---

# Project Description

This document analyzes krill biodiversity data from the Southern Ocean Continuous Plankton Recorder (SO-CPR) program. The analysis includes:

- Biodiversity indices calculation (Richness, Shannon, Simpson, ES50)
- Spatial visualization of biodiversity patterns
- Taxonomic occurrence tables
- Temporal abundance trends
- Latitudinal abundance patterns
- Species-specific distribution mapping

The analysis is modular and filters can be applied prior to fit specific visualization needs


# 1. Environment Setup
## 1.1 Package Installation and Loading
We first install and load all required packages. The code checks for missing packages and installs them automatically.

```{r Install and Load libraries, include=TRUE}
# document settings
knitr::opts_chunk$set(
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	include = FALSE
)
# Function to check and install CRAN packages if not already installed
install_if_missing <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
    }
    library(pkg, character.only = TRUE)
  }
}

# Install and load CRAN packages
cran_packages <- c("tidyverse", "sf", "leaflet", "viridis", "reactable", 
                   "ggplot2", "worrms", "dplyr", "tidyr", "stringr", "mgcv", "fields",
                   "rnaturalearth", "rnaturalearthdata")
install_if_missing(cran_packages)

# Install and load GitHub packages if not already installed
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("h3", quietly = TRUE)) remotes::install_github("crazycapivara/h3-r")
if (!requireNamespace("SOmap", quietly = TRUE)) remotes::install_github("AustralianAntarcticDivision/SOmap")

# Load GitHub packages
library(h3)
library(SOmap)

# Load additional libraries
library(blueant)
```

## 1.2 Analysis Parameters
Set global parameters for the biodiversity analysis. These can be modified to change the analysis scope.

```{r Analysis Parameters, include = TRUE}
# Choose working directory and location of taxon annotations file created with SO_CPR_AnnotateSpeciesTaxonomy.R script
workdir = "C:/yourdir/SO-CPR_RawData_BlueAnt/"
taxon.annotations = "C:/yourdir/SO-CPR_ID_TaxonAnnotations.txt"

# Spatial resolution for H3 hexagons to plot the spatial biodiversity (1-15, higher = finer)
res = 3

# Temporal and ship code filters (NULL = no filter)
Month_select = NULL       # c("January", "February", ...)
Ship_code_select = NULL   # Ship codes to include
Year_select = NULL       # Years to include
Season_select = NULL     # Seasons to include

# Taxonomic filters (NULL = no filter)
Taxon = "Rotaliida" # Regex pattern to filter taxonomy

# Define the bounding box ranges for Longitude and Latitude 
# Disclaimer: The bounding box is applied as an extra filter to generate a table with abundances of every species of/within the selected taxon and temporal trends. Set the bounding box values to NULL in order to avoid spatial filtering.
xmin <- NULL   
xmax <- NULL   
ymin <- NULL   
ymax <- NULL   
```

# 2. Data Preparation
## 2.1 Load CPR Krill Data
Data is loaded from the BlueAnt repository containing SO-CPR data.

```{r Load CPR Krill Data, include = TRUE}
## Read CPR Krill data using BlueAnt ####
# Configure data folder
cpr_krill_folder <- workdir
cpr_krill <- bb_config(local_file_root = cpr_krill_folder)

# Select and add CPR krill dataset from BlueAnt
SO_plankton <- sources("Southern Ocean Continuous Plankton Recorder")
cpr_krill <- cpr_krill %>% bb_add(SO_plankton)

# Synchronize/download the dataset
status <- bb_sync(cpr_krill)

# Load the CPR data
myfiles <- status$files[[1]]
cpr_data_input <- read.csv(myfiles$file[grepl("AADC", myfiles$file)])

# Column names of metadata and plankton species
colnames_metadata <- c("Tow_Number", "Ship_Code", "Time", "Date", 
                       "Month", "Year", "Season", "Latitude", 
                       "Longitude", "Segment_No.", "Segment_Length", 
                       "Total.abundance", "Phytoplankton_Colour_Index", 
                       "Fluorescence", "Salinity", "Water_Temperature", 
                       "Photosynthetically_Active_Radiation")
colnames_species <- setdiff(colnames(cpr_data_input), colnames_metadata)
```

## 2.2 Load Taxonomic Metadata
Load hierarchical taxonomic classification for species IDs. The classification dataset was made with the SO_CPR_AnnotateSpeciesTaxonomy.R script. 

```{r Load Taxonomic Metadata, include = TRUE}
## Read taxonomic metadata for every ID ####
taxonomy <- read.table(
  file = taxon.annotations,
  sep = "\t", 
  header = TRUE
)
```

## 2.3 Apply Filters
Filter data based on specified parameters.

```{r Apply Filters, include = TRUE}
## Apply filters only if the arguments are not NULL ####

# Filter based on taxonomy
taxonomy_keep <- taxonomy

if (!is.null(Taxon)) {
  taxonomy_keep <- taxonomy_keep %>%
    filter(
      grepl(Taxon, species, ignore.case = TRUE) |
        grepl(Taxon, genus, ignore.case = TRUE) |
        grepl(Taxon, family, ignore.case = TRUE) |
        grepl(Taxon, suborder, ignore.case = TRUE) |
        grepl(Taxon, order, ignore.case = TRUE) |
        grepl(Taxon, subclass, ignore.case = TRUE) |
        grepl(Taxon, class, ignore.case = TRUE) |
        grepl(Taxon, subphylum, ignore.case = TRUE) |
        grepl(Taxon, phylum, ignore.case = TRUE) |
        grepl(Taxon, infrakingdom, ignore.case = TRUE) |
        grepl(Taxon, subkingdom, ignore.case = TRUE) |
        grepl(Taxon, kingdom, ignore.case = TRUE)
    )
}

selected_columns_IDs <- taxonomy_keep$ID

# Subset cpr_data_input using the selected taxon
cpr_data_taxons_filtered <- cpr_data_input %>%
  dplyr::select(all_of(c(selected_columns_IDs, colnames_metadata)))

# Filter the data with additional arguments (month, ship code,...)
cpr_data_filtered <- cpr_data_taxons_filtered %>%
  filter(
    if (!is.null(Month_select)) Month %in% Month_select else TRUE
  ) %>%
  filter(
    if (!is.null(Ship_code_select)) Ship_code %in% Ship_code_select else TRUE
  ) %>%
  filter(
    if (!is.null(Year_select)) Year %in% Year_select else TRUE
  ) %>%
  filter(
    if (!is.null(Season_select)) Season %in% Season_select else TRUE
  )

# Add an extra column "Taxon.abundance" to represent the abundance of the selected taxon for each row.
cpr_data_taxons_filtered <- cpr_data_taxons_filtered %>%
  mutate(Taxon.abundance = rowSums(dplyr::select(., all_of(selected_columns_IDs)), na.rm = TRUE)) %>%
  filter(Taxon.abundance > 0)

# Filter data based on bounding box
if (!is.null(xmin) & !is.null(xmax) & !is.null(ymin) & !is.null(ymax)) {
  cpr_data_bbox <- cpr_data_taxons_filtered %>%
    filter(Longitude >= xmin & Longitude <= xmax,
           Latitude >= ymin & Latitude <= ymax)
} else {
  cpr_data_bbox <- cpr_data_taxons_filtered
}

# Show filtered data dimensions
dim(cpr_data_bbox)
```

# 3. Biodiversity Analysis
## 3.1 Calculate Biodiversity Indices
We calculate several biodiversity metrics at the H3 hexagon level:

- Species Richness
- Shannon Diversity Index
- Simpson Diversity Index
- ES50 (Hurlbert's Expected Species)

```{r Calculate Biodiversity Indices, echo=FALSE}
# Prepare data with H3 spatial indexing
resolution = res
cpr_data <- cpr_data_filtered %>% 
  mutate(Longitude = Longitude, Latitude = Latitude) %>%
  st_as_sf(coords = c("Longitude", "Latitude")) %>%
  mutate(cell = geo_to_h3(., res = resolution))
cpr_data$Longitude <- cpr_data_filtered$Longitude
cpr_data$Latitude <- cpr_data_filtered$Latitude

# Function to calculate Shannon Diversity
shannon_diversity <- function(species_data) {
  species_abundances <- species_data[species_data > 0]
  if (length(species_abundances) == 0) return(0)
  p_i <- species_abundances / sum(species_abundances)
  H <- -sum(p_i * log(p_i))
  return(H)
}

# Function to calculate Simpson Diversity
simpson_diversity <- function(species_data) {
  species_abundances <- species_data[species_data > 0]
  if (length(species_abundances) == 0) return(0)
  p_i <- species_abundances / sum(species_abundances)
  return(1 - sum(p_i^2))
}

# Function to calculate the ES50 (Hurlbert's index)
calculate_es50 <- function(species_counts) {
  n <- sum(species_counts)
  if (n < 50) return(NA)
  
  species_present <- species_counts[species_counts > 0]
  
  es_values <- sapply(species_present, function(ni) {
    if ((n - ni) >= 50) {
      return(1 - exp(lgamma(n - ni + 1) + lgamma(n - 50 + 1) - lgamma(n - ni - 50 + 1) - lgamma(n + 1)))
    } else {
      return(1)
    }
  })
  
  return(sum(es_values, na.rm = TRUE))
}

# Calculate biodiversity metrics per hexagon cell
cpr_richness <- cpr_data %>%
  group_by(cell) %>%
  summarise(
    survey_effort = n(),
    across(all_of(selected_columns_IDs), ~ sum(. >= 1))
  ) %>%
  na.omit() %>%
  data.frame()

# Add diversity metrics
cpr_richness$shannon_diversity <- apply(cpr_richness[, which(names(cpr_richness) %in% selected_columns_IDs)], 1, shannon_diversity)
cpr_richness$simpson_diversity <- apply(cpr_richness[, which(names(cpr_richness) %in% selected_columns_IDs)], 1, simpson_diversity)
cpr_richness$richness <- apply(cpr_richness[, which(names(cpr_richness) %in% selected_columns_IDs)], 1, function(x) sum(x > 0))
cpr_richness <- cpr_richness %>% 
  mutate(shannon_diversity_corr = shannon_diversity/survey_effort,
         simpson_diversity_corr = simpson_diversity/survey_effort,
         richness_corr = richness/survey_effort)
cpr_richness$ES50 <- apply(cpr_richness[, which(names(cpr_richness) %in% selected_columns_IDs)], 1, calculate_es50)
```

# 4. Visualization
## 4.1 Spatial Biodiversity Patterns
Visualize biodiversity metrics across the Southern Ocean.

```{r Spatial Biodiversity Patterns, include = TRUE}
# Convert H3 cells to polygons for visualization
polygons_richness <- h3_to_geo_boundary_sf(cpr_richness$cell) %>%
  dplyr::mutate(cell = cpr_richness$cell,
                survey_effort = cpr_richness$survey_effort,
                richness = cpr_richness$richness,
                shannon_diversity = cpr_richness$shannon_diversity,
                simpson_diversity = cpr_richness$simpson_diversity,
                richness_corr = cpr_richness$richness_corr,
                shannon_diversity_corr = cpr_richness$shannon_diversity_corr,
                simpson_diversity_corr = cpr_richness$simpson_diversity_corr,
                ES50 = cpr_richness$ES50)

# Reproject to Antarctic South Polar Stereographic (EPSG 3031)
polygons_richness <- st_transform(polygons_richness, crs = 3031)

# Define richness type and gradient size
Richness_type <- "ES50"
gradient.size <- 100

# Generate colors based on values
richness_values <- polygons_richness[[Richness_type]]
color_palette <- viridis(gradient.size)
color_breaks <- cut(as.numeric(richness_values), gradient.size, labels = FALSE)

# Plot SOmap as the base layer
SOmap(trim = -35, graticules = FALSE, ice = TRUE)

# Overlay polygons with color mapping
plot(st_geometry(polygons_richness), 
     col = color_palette[color_breaks], 
     add = TRUE, 
     border = "gray")

# Add legend
legend_titles <- list(
  "richness" = "Species Richness",
  "shannon_diversity" = "Shannon Diversity",
  "simpson_diversity" = "Simpson Diversity",
  "richness_corr" = "Species Richness (corrected)",
  "shannon_diversity_corr" = "Shannon Diversity (corrected)",
  "simpson_diversity_corr" = "Simpson Diversity (corrected)",
  "ES50" = "ES50 (Hurlbert index)"
)
legend_label <- legend_titles[[Richness_type]]                      
image.plot(
  legend.only = TRUE,
  col = color_palette,
  zlim = range(richness_values, na.rm = TRUE),
  legend.lab = legend_label
)
```

## 4.2 Species Occurrence Table
Interactive table showing species occurrences in a specified bounding box.

```{r Species Occurrence Table, include = TRUE}

# Create species occurrence table
cpr_data_bbox_ID <- cpr_data_bbox[, which(colnames(cpr_data_bbox) %in% c(selected_columns_IDs))]
species_presence <- colSums(cpr_data_bbox_ID) %>% na.omit() %>% .[.>0] %>% sort(decreasing = TRUE)
species_table <- data.frame(Species = names(species_presence), Abundance = as.numeric(species_presence))

# Interactive table
reactable(species_table, 
          columns = list(
            Species = colDef(name = "Species", width = 150),
            Abundance = colDef(name = "Abundance", format = colFormat(separators = TRUE))
          ),
          bordered = TRUE, striped = TRUE, highlight = TRUE,
          defaultPageSize = 10)
```

## 4.3 Temporal Trends
Plot annual abundance patterns for selected taxa in the bounding box.

```{r Temporal Trends, include = TRUE}
# Prepare yearly data
cpr_data_bbox_ID.Year <- cpr_data_bbox[, which(colnames(cpr_data_bbox) %in% c(selected_columns_IDs, "Year"))]

year_counts <- cpr_data_bbox_ID.Year %>%
  count(Year, name = "Yearly_Surveys")

species_presence_yearly <- cpr_data_bbox_ID.Year %>%
  group_by(Year) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
  pivot_longer(-Year, names_to = "Species", values_to = "Abundance") %>%
  filter(Abundance > 0) %>%
  arrange(Year, desc(Abundance)) %>%
  left_join(year_counts, by = "Year") %>%
  mutate(Corrected_Abundance = Abundance / Yearly_Surveys)

total_abundance_yearly <- species_presence_yearly %>%
  group_by(Year) %>%
  summarise(Total_Abundance = sum(Abundance),
            Yearly_Surveys = sum(Yearly_Surveys),
            Total_Corrected_Abundance = sum(Corrected_Abundance))

# Plot corrected abundance per species
ggplot(species_presence_yearly, aes(x = Year, y = Corrected_Abundance, group = Species, color = Species)) +
  geom_line() +
  geom_point() +
  labs(title = "Annual corrected abundance of species",
       x = "Year",
       y = "Abundance (Corrected for Sampling Effort)",
       color = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Plot total abundance trends
ggplot(total_abundance_yearly, aes(x = Year)) +
  geom_line(aes(y = Total_Abundance, color = "Uncorrected"), linewidth = 1) +
  geom_point(aes(y = Total_Abundance, color = "Uncorrected"), size = 2) +
  geom_line(aes(y = Total_Corrected_Abundance, color = "Corrected"), linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = Total_Corrected_Abundance, color = "Corrected"), size = 2) +
  geom_line(aes(y = Yearly_Surveys, color = "lightgrey")) +
  scale_color_manual(values = c("Uncorrected" = "blue", "Corrected" = "red")) +
  labs(title = "Total annual abundance (Corrected vs. Uncorrected)",
       x = "Year",
       y = "Total Abundance",
       color = "Abundance Type") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 4.4 Monthly Abundance Patterns
Boxplot showing monthly variation in total abundance in the bounding box.

```{r Monthly Abundance Patterns, include = TRUE}
# Ensure Month is ordered correctly
cpr_data_bbox$Month <- factor(cpr_data_bbox$Month, 
                                        levels = month.name)
# Calculate taxon total abundance for each row
cpr_data_bbox <- cpr_data_bbox %>%
  mutate(Taxon.abundance = rowSums(dplyr::select(., all_of(selected_columns_IDs)), na.rm = TRUE)) %>%
  filter(Taxon.abundance > 0)

ggplot(cpr_data_bbox, aes(x = Month, y = Taxon.abundance)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "darkgrey", outlier.size = 0.5) +
  labs(title = "Total abundance per Month",
       x = "Month",
       y = "Total abundance per five mile segment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, max(cpr_data_bbox$Taxon.abundance) * 1.05)
```

## 4.5 Latitudinal Patterns
GAM smooth of abundance along latitude.

```{r Latitudinal Patterns, include = TRUE}

# Fit GAM model
gam_model <- gam(Taxon.abundance ~ s(Latitude), data = cpr_data_taxons_filtered)

# Plot with GAM smooth
ggplot(cpr_data_taxons_filtered, aes(x = Latitude, y = Taxon.abundance)) +
  geom_point(color = "blue", size = 2, alpha = 0.7) + 
  geom_smooth(method = "gam", formula = y ~ s(x), color = "red", se = TRUE, linetype = "dashed") +
  labs(title = "Latitudinal occurence",
       subtitle = "Using a Generalized Additive Model",
       x = "Latitude",
       y = "Total abundance per five mile segment") +
  theme_minimal(base_size = 15) +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
        plot.subtitle = element_text(hjust = 0.5, size = 15)) +  
  ylim(0, max(cpr_data_bbox$Taxon.abundance) * 1.05)
```

## 4.6 Species Occurrence Map
Spatial distribution of a selected species.

```{r Species Distribution Map, include = TRUE}

# Prepare presence data
presence_data <- cpr_data_taxons_filtered %>%
  mutate(
    Longitude = as.numeric(Longitude),
    Latitude = as.numeric(Latitude)
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = 3031)

# Prepare background data
background_data <- cpr_data_input %>%
  dplyr::select(Longitude, Latitude) %>%
  mutate(
    Longitude = as.numeric(Longitude),
    Latitude = as.numeric(Latitude)
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = 3031)

# Get land data
land <- ne_countries(scale = "medium", returnclass = "sf") %>%
  st_transform(crs = 3031)

# Plot distribution
ggplot() +
  geom_sf(data = land, fill = "grey90", color = "black", size = 0.3) +
  geom_sf(data = background_data, color = "grey", size = 0.5, alpha = 0.2) +
  geom_sf(data = presence_data, aes(size = Taxon.abundance), color = "blue", alpha = 0.1) +
  scale_size_continuous(name = "Abundance", range = c(0.5, 4)) +
  coord_sf(crs = st_crs(3031), ylim = c(-4500000, 4500000), xlim = c(-4500000, 4500000)) +
  theme_minimal() +
  labs(
    title = paste("Presence of", Taxon),
    subtitle = "Grey lines represent sampling trajectories",
    x = "", y = ""
  ) +
  theme(legend.position = "right")
```

# 5. Session Information

```{r Session Information, include = TRUE}
sessionInfo()
```
